signalHsmm - a novel semi-Markov model of eukaryotic signal peptides

### 2015-06-15 paper.tex
added signal peptide figure

\title{signalHsmm - a novel semi-Markov model of signal peptides}

@@

\begin{abstract}
The proper localization of proteins in a cell is essential to perform their desired function. Information about the protein destination is included within the very protein in the form of short peptides called targeting signals. Ones of them are signal peptides, diverse N-terminal sequences, which are responsible for targeting of proteins to endomembrane system and their export outside the cell. Proteins equipped with signal peptides constitute a substantial fraction of the whole proteome and play crucial roles in metabolism, maintenance of tissue structure, immune response and regulation of other organismal functions. Moreover, the transport of proteins through the endomembrane system is important for their correct folding and posttranslational modifications. A common model of classical signal peptides assumes that they start with a positively charged n-region, followed by a hydrophobic h-region and a c-region ended with a cleavage site recognised by a signal peptidase.

However, our studies of many protein sequences representing the wide range of diversified taxonomic organisms indicate a variability of signal peptides. Therefore, the main goal of the study is to design a new probabilistic model for signal peptides, which will include knowledge about their organization, amino acid composition and variation.

The proposed model will be based on hidden semi-Markov models (HSMMs) and use intrinsic knowledge about signal peptides. The big advantage of the algorithm is a possibility to incorporate unique regions in the architecture of signal peptides. Therefore, HSMMs can be useful in recognition of atypical and artificial signal peptides. An aggregation of amino acids into physicochemical groups using the neural gas algorithm with k-fold cross-validation will reduce dimensionality of the problem and enable to learn the algorithm even on smaller data sets. The reduction will allow to use more complex probabilistic models and models specialized in the detection of signal peptides in different taxonomic groups of organisms, which improve the efficiency of the whole predictor. The n-gram and position-specific weight matrix methods will add more details to the probabilistic model.

@@

Your introduction goes here.
Some examples of commonly used {\LaTeX} commands and features are listed below, to help you get started.

All \textbf{accepted} submissions will be published as a \textbf{preprint collection} with \textbf{PeerJ} (\url{http://www.peerj.com}).
It will be clearly visible that the work
\begin{enumerate}
\item is a preprint (although reviewed and accepted by the GCB program committee)
\item has been presented at GCB'15
\end{enumerate}
by including a special GCB logo in the final version

If you decide to submit a full version after GCB to PeerJ or to PeerJ Computer Science, the procedure is very easy.
Of course, you are free to submit your paper to any other journal that allows existing preprints.

\subsection*{About PeerJ}

PeerJ is an award-winning open access publisher covering both computer science and the biological and medical sciences. 
PeerJ provides authors with three publication venues: PeerJ and PeerJ Computer Science (peer-reviewed academic journals) and PeerJ PrePrints (a 'pre-print server'). See \url{https://peerj.com/about/publications/} for more information.

The PeerJ model allows an author to publish articles in their peer-reviewed journal via the purchase of a lifetime Publication Plan. Prices start from just \$99 (a one-off payment) which entitles an author to the lifetime ability to publish 1 article per year for free.
The fee applies to all authors of an article, which encourages a reasonable number of co-authors and is generally less than for other open-access publishers.

\section*{Some \LaTeX{} Examples}
\label{sec:examples}

Use section and subsection commands to organize your document. \LaTeX{} handles all the formatting and numbering automatically. Use ref and label commands for cross-references.

\subsection*{Figures and Tables}

Use the table environment and the tabular command for basic tables --- see Table~\ref{tab:widgets}, for example.

To include a figure in your document, use the figure environment and the includegraphics command as in the code for Figure~\ref{fig:uaruhr}.

\begin{figure}[ht]\centering
\includegraphics[width=0.6\textwidth]{uaruhr}
\caption{An example image.}
\label{fig:uaruhr}
\end{figure}

\begin{table}[ht]
\centering
\begin{tabular}{l|r}
Item & Quantity \\\hline
Widgets & 42 \\
Gadgets & 13
\end{tabular}
\caption{\label{tab:widgets}An example table.}
\end{table}

\subsection*{Citations}

LaTeX formats citations and references automatically using the bibliography records in your .bib file.
Use the \verb|\cite| command for an inline citation, like

\subsection*{Mathematics}

\LaTeX{} is great at typesetting mathematics. Let $X_1, X_2, \ldots, X_n$ be a sequence of independent and identically distributed random variables with $\text{E}[X_i] = \mu$ and $\text{Var}[X_i] = \sigma^2 < \infty$, and let
$$S_n = \frac{X_1 + X_2 + \cdots + X_n}{n}
      = \frac{1}{n}\sum_{i}^{n} X_i$$
denote their mean. Then as $n$ approaches infinity, the random variables $\sqrt{n}(S_n - \mu)$ converge in distribution to a normal $\mathcal{N}(0, \sigma^2)$.

\subsection*{Lists}

You can make lists with automatic numbering \dots

\begin{enumerate}[noitemsep] 
\item Like this,
\item and like this.
\end{enumerate}
\dots or bullet points \dots
\begin{itemize}[noitemsep] 
\item Like this,
\item and like this.
\end{itemize}
\dots or with words and descriptions \dots
\begin{description}
\item[Word] Definition
\item[Concept] Explanation
\item[Idea] Text
\end{description}

@@

@@

We cluster amino acids using several criteria relevant for the architecture of signal peptide: hydrophobicity, frequency in alpha-helices, polarity and size. High values of hydrophobicity are required in the h-region, the core of signal peptides. Alpha-helix, the secondary structure of this region, is probably induced by the positively charged n-region. Polarity as well as size are important especially in the cleavage site~\citep{1994palzkillselection}.

@@

We selected 13 properties from AAIndex database (aaindexcitation) (see Table~\ref{tab:aaprop}). Each property was attributed to a single criterion. Considering all combinations of single properties, we created 96 possible clusterings of amino acids using Euclidean distance and Ward's method.

\subsection*{Subsection}

Here is an interesting equation that may be helpful in some situations:
\begin{equation}
\cos^3 \theta =\frac{1}{4}\cos\theta+\frac{3}{4}\cos 3\theta
\label{eq:refname2}
\end{equation}

\paragraph{Paragraph}
Nothing to see here. Move on.

\paragraph{Paragraph}
Really.
See Figure~\ref{fig:results} for more interesting results.

\begin{figure}[ht]\centering
\includegraphics[width=0.6\textwidth]{results}
\caption{Can you guess which function this is?}
\label{fig:results}
\end{figure}

\section*{Results and Discussion}

@@

>>>>

\title{signalHsmm - a novel semi-Markov model of eukaryotic signal peptides}

@@

\begin{abstract}
The proper localization of proteins in a cell is essential to maintain their desired function. Information about the protein destination is included within the very protein in the form of short peptides called targeting signals. Ones of them are signal peptides, diverse N-terminal sequences, which are responsible for targeting of proteins to endomembrane system and their export outside the cell. Proteins equipped with signal peptides play crucial roles in metabolism, maintenance of tissue structure, immune response and regulation of other organismal functions. Moreover, the transport of proteins through the endomembrane system is important for their correct folding and posttranslational modifications.

A common model of classical signal peptides assumes that they start with a positively charged n-region, followed by a hydrophobic h-region and a c-region ended with a cleavage site recognised by a signal peptidase. However, our studies of many protein sequences representing the wide range of diversified taxonomic organisms indicate a variability of signal peptides. Therefore, we designed a new, more univeral probabilistic model for eukaryotic signal peptides, which includes knowledge about their organisation, amino acid composition and variation.

The proposed model is based on hidden semi-Markov models (HSMMs) and use intrinsic knowledge about signal peptides. The big advantage of the algorithm is its extensibility. Using the k-mers we point how the general model can be attuned to yield not only better results, but also more information about signal peptides.

@@

Proteins of eukaryotes are encoded in nuclear genomes and are synthetized in ribosomes located in the cytosol or bounded by the endoplasmic reticulum. After the translation process, proteins have to be targeted to specific subcellular compartments or exported outside the cell to the extracellular environment. The proper localization of proteins is essential to perform their desired function. Information about protein destination is included within the very protein in the form of short peptides or stretches of amino acid residues called targeting or sorting signals. Ones of them are signal peptides, which are located at the N-terminus of proteins.

Signal peptides are responsible for targeting of proteins to endomembrane system, which includes endoplasmic reticulum, Golgi apparatus and endosomes. Such proteins can stay inside of these compartments, or can be inserted into cellular membranes or exported outside the cell. Proteins equipped with signal peptides constitute a substantial fraction of the whole proteome. They play crucial roles in metabolism ($\beta$ galactosidase, pepsins)~\citep{1991hofmannmutations}, maintenance of tissue structure (collagen)~\citep{2001chanaberrant}, immune response (interferons, interleukins)~\citep{2005zhangalteration} and regulation of other organismal functions (prolactin, glucagon)~\citep{2010huangrole}. Moreover, passing proteins through the endomembrane system is important for their correct folding and posttranslational modification such as glycosylation and phosphorylation.

Although many experimental methods determining the subcellular localization of proteins were devised, they are time consuming and laborious. Therefore, the development of new approaches in the field of computational biology and bioinformatics is desirable. They are not only a good support, complement or alternative for the experimental methods but also enable to understand rules encoding information about protein targeting, which is contained in the predicted targeting signals. However, many of the present predictors disregard full biological information carrying by signal peptides. Moreover, they often use orthogonal amino acids encodings, which greatly increase the model complexity without preserving physiochemical properties of residues. The current signal predicting software is mainly based on sequences of frequently studied species, such as yeast, angiosperms and mammals. However, our analysis of the large number of sequences representing diverse taxonomical groups indicates a great variability of signal peptides. The mentioned limitations result in inflexible models, which inefficiently extract data and henceforth cannot be used to model signal peptides restricted to smaller taxonomic groups and atypical.

Taking into account the importance of proteins with signal peptides, the aim of the project became a design for a new probabilistic model for signal peptide, which will include knowledge about their organization, amino acid composition and variation. The model will be based on hidden semi-Markov models. It will be created for different taxonomic groups of organisms, validated through a prediction algorithm and incorporated into a user-friendly software. The program will be also compared with other signal peptide predictors and competence area of such software will be determined. In the project, we are also planning to analyse differences in amino acid composition of signal peptides in different groups of organisms and recognise various types of signal peptides according to their distinctive features.

@@

\begin{figure}[ht]\centering
\includegraphics[width=0.5\textwidth]{SP.png}
\caption{The structure of signal peptide.}
\label{fig:sparch}
\end{figure}

@@

Amino acids were clustered using several criteria relevant for the architecture of signal peptide: hydrophobicity, frequency in alpha-helices, polarity and size. High values of hydrophobicity are required in the h-region, the core of signal peptides. Alpha-helix, the secondary structure of this region, is probably induced by the positively charged n-region. Polarity as well as size are important especially in the cleavage site~\citep{1994palzkillselection}.

@@

We selected 13 properties from AAIndex database (aaindexcitation) (see Table~\ref{tab:aaprop}), each attributed to a single criterion. Considering all combinations of single properties for every criterion, we created 96 possible clusterings of amino acids using Euclidean distance and Ward's method.

To compare encodings we performed a 5-fold cross-validation training a new instance of signalHsmm on every encoding. We created balanced data sets by subsampling a number of proteins without a signal peptide equal to the number of proteins with signal peptide. The cross-validation was repeated XXX times, to ensure that every negative protein was tested at least once with XXX probability.

\subsection*{Hidden semi-Markov model}

Tu troche matematyki o ukrytych modelach Markova.

\section*{Results}

@@

\section*{Discussion}
